Ce que j'ai compris jusqu'à maintenant\\Début de rapport ?

### 2025-04-04 article_bihom_v2.tex
v1.5

@@

		The major "issue" of Gröbner basis algorithms is the reductions to 0, because the size of the matrices are huge, multiple millions for practical examples,
		reductions to 0 can cost a lot during the linear algebra process. F5 \cite{F02} was a major improvement because it provided a criterion that avoided reduction to 0 during the linear algebra part for all regular sequences.
		Bardet showed in \cite{Bardet04} that F5 criterion removed reductions to 0 for semi-regular sequences until the degree $D$ which is the degree of regularity of $I$ which is the ideal generated by the polynomials of the system.

		\begin{theorem}[Theorem 3.2.10 in \cite{Bardet04}]
			Let $f_1,...,f_m$ a homogeneous sequence, such that $\langle f_1,...,f_m \rangle$ is 0-dimensional and
			$\prec$ an admissible graded monomial ordering. We have,
			\begin{itemize}
				\item[-] If the sequence is semi-regular, then no reductions to 0 are performed during the F5-matrix algorithm until the degree of regularity $D - 1$.
				\item[-] If there is no reductions to 0 during the F5-matrix algorithm until the degree $D - 1$, and if the matrix of degree D is full rank and is the first matrix to have more rows than columns, then the sequence is semi-regular and it's index of regularity is $H(I) = D$.
			\end{itemize}
		\end{theorem}

@@

@@

		\subsubsection{Jacobian matrices and syzygies}
		First we will need to introduce some important notations.
		\begin{itemize}
			\item[-] $R = \mathbb{K}[x_1,...,x_{n_x},y_1,...,y_{n_y}]$
			\item[-] Let $\mathbb{K}$ be a field and $f_1,...,f_m \in R$ be a bilinear polynomial. We denote by $F_i$ the polynomial sequence $(f_1,...,f_i)$ and $I_i$ the ideal spanned by this sequence $\langle F_i \rangle$.
			\item[-] $jac_{x}(F_{i-1})$ (resp. $jac_{y}(F_{i-1})$) is the jacobian matrix with respect to the two subsets of variables $x_1,...,x_{n_x}$ (resp. $y_1,...,y_{n_y}$).
			\item[-] Let $M$ be a $l \times c$ matrix with $l > c$. We call maximal minors of $M$ the determinants of the $c \times c$ sub-matrices of $M$. 
			\item[-] $I_{i-1} : f_i$ is the ideal spanned by $I_{i-1}$ and by the maximal minors of $jac_{x}(F_{i-1})$ if ($i > n_y$) and $jac_{y}(F_{i-1})$ if ($i > n_x$).
		\end{itemize}
		We will give important results of this article before discussing them further.

@@

		The following algorithms will give us the signatures of reductions to 0 during F5-matrix algorithm.

@@

		 \begin{proposition}
		 	The complexity of algorithm \ref{alg:crit_calc} is
		 	$$
		 	\mathcal{O}\left(\left(\sum_{i = n_x}^{m} \binom{i}{n_x}n_x^{\omega} + \binom{2n_x + 1}{n_x}^\omega\right) + \left(\sum_{i = n_y}^{m} \binom{i}{n_y}n_y^{\omega} + \binom{2n_y + 1}{n_y}^\omega\right)\right)
		 	$$
		 \end{proposition}

@@

		 If this case arises, the only computations that would be done are for extended F5 criterion. The complexity for F5-Matrix algorithm with only the Frobenius criterion is 
		 $$
		 \mathcal{O}\left(\binom{n + 4}{4}^\omega\right)
		 $$
		 If we only take into consideration the big O notation and not the constants, which is not realistic at all for efficient implementations to really break a cryptographic scheme, we get
		 \begin{equation}
		 Adv(n, m) = 2\left(\sum_{i = n/2}^{m}\binom{i}{n/2}\left(\frac{n}{2}\right)^\omega + \binom{n + 1}{n/2}^\omega\right) - \binom{n + 4}{4}^\omega
		 \label{equation_complex}
		 \end{equation}

>>>>

\newcommand{\Mac}[3]{{Mac_{#1, #2}(\mathcal{#3})}}
\newcommand{\Mact}[3]{{\widetilde{Mac}_{#1, #2}(\mathcal{#3})}}

@@

		In \cite{Lazard83}, Lazard introduced the connection between linear algebra and Gröbner basis theory via the usage of Macaulay matrices which is defined as follows.

		\begin{definition}
			Let $\prec$ be an admissible monomial ordering on a given ring $R = \mathbb{K}[x_1,\dots,x_n]$ where $\mathbb{K}$ is a field. Given a sequence of homogeneous (resp. affine) polynomials $\mathcal{F} = (f_1,\dots,f_m) \in R$, we associate to it the Macaulay matrix of degree $D$ (resp. $\leq D$), denoted by $\Mac{D}{m}{\mathcal{F}}$ (resp. $\Mac{\leq D}{m}{\mathcal{F}}$), and defined as follows: the columns of the matrix are indexed by the monomials of degree $D$ (resp. $\leq D$) sorted in decreasing order from the left to right using the defined monomial ordering. Each row of the matrix is labeled by a tag, also called signature, ($u, f_i$) where u is a monomial in $R$ and $f_i \in \mathcal{F}$ such that $deg(uf_i) = D$ (resp. $deg(uf_i) \leq D$), and the polynomial $uf_i$ is written as vector of coefficients of monomials on the row. We denote by $\Mact{D}{m}{F}$ (resp. $\Mact{\leq D}{m}{F}$) the Macaulay matrix on which we apply the row echelon form without swapping columns to preserve the monomial ordering.
		\end{definition}

		The major "issue" of Gröbner basis algorithms is the reductions to 0 in the Macaulay matrix associated to a polynomial sequence $\Mac{D}{m}{F}$, because the size of the matrices are huge, multiple millions for practical examples,
		reductions to 0 can cost a lot during the linear algebra process. F5 \cite{F02} was a major improvement because it provided a criterion that avoided reduction to 0 during the linear algebra part for all regular sequences.

		We will introduce the F5 criterions and the notions of regularity and semi-regularity in the following.

		\begin{proposition}[General criterion \cite{F02}]
			\label{F5Crit}
			Let $d_i$ be the total degree of $f_i$.
			$\forall j < m$, if a row of signature ($u, f_j$) in the matrix $\Mact{D-d_j}{m-1}{F}$ has as leading term $t'$, then the row ($t', f_m$) in $\Mac{D}{m}{F}$ (resp. $\Mac{\leq D}{m}{F}$) will reduce to 0 i.e. is a linear combination of the predecessors.
		\end{proposition}

		\begin{proposition}[Frobenius criterion \cite{F02}]
			\label{Frob}
			If a row of signature ($t, f_m$) in $\Mact{D - d_m}{m}{F}$ has as leading term $t'$, then the row ($t', f_m$) in $\Mac{D}{m}{F}$ (resp. $\Mac{\leq D}{m}{F}$) will reduce to 0 i.e. is a linear combination of the predecessors.
		\end{proposition}

		Bardet showed in \cite{Bardet04} that F5 criterion removed reductions to 0 for semi-regular sequences until the degree $D$ which is the degree of regularity of $I$ which is the ideal generated by the polynomials of the system.

		\begin{theorem}[Theorem 3.2.10 in \cite{Bardet04}]
			Let $f_1,...,f_m$ a homogeneous sequence, such that $\langle f_1,...,f_m \rangle$ is 0-dimensional and
			$\prec$ an admissible graded monomial ordering. We have,
			\begin{itemize}
				\item[-] If the sequence is semi-regular, then no reductions to 0 are performed during the F5-matrix algorithm until the degree of regularity $D - 1$.
				\item[-] If there are no reductions to 0 during the F5-matrix algorithm until the degree $D - 1$, and if the matrix of degree D is full rank and is the first matrix to have more rows than columns, then the sequence is semi-regular and its index of regularity is $H(I) = D$.
			\end{itemize}
		\end{theorem}

@@

		\begin{definition}
			Let $E, F, G$ be vector spaces on the same base field $\mathbb{K}$. 
			Let $f: E \times F \leftarrow G$ an application. We say that $f$ is bilinear if
			$$
				\forall (x, x') \in E^2 , \forall (y, y') \in F^2 , \forall \lambda \in \mathbb{K}
			$$
			$$
				f(x + x',y) = f(x, y) + f(x', y)
			$$
			$$
				f(x, y + y') = f(x, y) + f(x, y')
			$$
			$$
				f(\lambda x, y) = \lambda f(x, y)
			$$
		\end{definition}

@@

		\subsubsection{Jacobian matrices and syzygies}
		First we will need to introduce some important notations.
		\begin{itemize}
			\item[-] Let $\mathbb{K}$ be a field. In our case in cryptography $\mathbb{K}$ is always finite.
			\item[-] $R = \mathbb{K}[x_1,...,x_{n_x},y_1,...,y_{n_y}]$
			\item[-] Let $f_1,...,f_m \in R$ be a bilinear polynomial. We denote by $F_i$ the polynomial sequence $(f_1,...,f_i)$ and $I_i$ the ideal spanned by this sequence $\langle F_i \rangle$.
			\item[-] $jac_{x}(F_{i-1})$ (resp. $jac_{y}(F_{i-1})$) is the jacobian matrix with respect to the two subsets of variables $x_1,...,x_{n_x}$ (resp. $y_1,...,y_{n_y}$).
			\item[-] Let $M$ be a $l \times c$ matrix with $l > c$. We call maximal minors of $M$ the determinants of the $c \times c$ sub-matrices of $M$. 
			\item[-] $I_{i-1} : f_i$ is the ideal spanned by $\{g \in R \mid gf_i \in I_{i-1}\}$.
		\end{itemize}
		We will give important results of this article before discussing them further.

@@

		The Algorithms \ref{alg:crit_calc} will give us the signatures of reductions to 0 during F5-matrix algorithm.

@@

		 \begin{proposition}
		 	The complexity of Algorithm \ref{alg:crit_calc} is
		 	$$
		 	\mathcal{O}\left(\left(\sum_{i = n_x}^{m} \binom{i}{n_x}n_x^{\omega} + \binom{2n_x + 1}{n_x}^\omega\right) + \left(\sum_{i = n_y}^{m} \binom{i}{n_y}n_y^{\omega} + \binom{2n_y + 1}{n_y}^\omega\right)\right)
		 	$$
		 \end{proposition}

@@

		 If this case arises, the only computations that would be done are for extended F5 criterion. The complexity for F5-Matrix algorithm with only the Frobenius criterion \ref{Frob} is 
		 $$
		 \mathcal{O}\left(\binom{n + 4}{4}^\omega\right)
		 $$
		 If we only take into consideration the big O notation and not the constants, which is not realistic at all for efficient implementations to really break a cryptographic scheme, we get
		 \begin{equation}
		 Adv(n, m) = 2\left(\sum_{i = n/2}^{m}\binom{i}{n/2}\left(\frac{n}{2}\right)^\omega + \binom{n + 1}{n/2}^\omega\right) - \binom{n + 4}{4}^\omega
		 \label{equation_complex}
		 \end{equation}
